---
title: "class05"
author: "Adithi (PID: A16653979)"
format: pdf
editor: visual
---

## Using GGPLOT

The ggplot2 package needs to be installed as it doesn't not come with R "out of the box"

We use the 'install.packages()' function to do this

To use ggplot, I need to load it up using 'library(ggplot2)'

```{r}
#install.packages (gglot2)
library(ggplot2) 
ggplot()

```

ALL ggplot figures have at least three things:

-   data (the stuff we want to plot)

-   aesthetic mapping (x,y)

-   geoms

    ```{r}
    ggplot (cars) + aes(x=speed, y=dist) + geom_point()
    ```

ggplot is not the only graphing system in R. There are lots of others. There is even "base R" graphics.

```{r}
plot(cars)
```

### **Creating Scatter Plots**

```{r}
ggplot(cars) + 
  aes(x=speed, y=dist) +geom_point() +geom_smooth() +
  labs(title = "Speed vs. Stopping Distances of Cars", x= "Speed (MPH)", y = "Stopping Distance (ft))", caption = "Dataset: 'cars'")+ theme_bw()

```

**Adding More Plot Aesthetics using aes()**

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt" 
genes <- read.delim(url) 
head(genes)
```

```{r}
nrow(genes) 
colnames(genes) 
ncol(genes) 
table(genes$State) 
round( table(genes$State)/nrow(genes) * 100, 2 )
```

```{r}
p <- ggplot (genes) + aes(x= Condition1, y = Condition2, col=State) + geom_point() 
p + scale_colour_manual( values = c("violet", "lightblue", "lightgreen"))+ 
  labs(title = "Gene Expression Changes upon Drug Treatment", x= "Control (No Drug)", y= "Drug Treatment")
```

#### **Going Further**

```{r}
#install.packages("gapminder") 
#install.packages ("dplyr")
library(gapminder) 
library(dplyr) 
gapminder_2007 <- gapminder %>% filter(year==2007) 
ggplot(gapminder_2007)+ aes(x=gdpPercap, y=lifeExp, color=continent, size = pop)+
  geom_point(alpha=0.5) 
ggplot(gapminder_2007)+ aes(x=gdpPercap, y=lifeExp, color = pop) +
geom_point(alpha=0.8) 

```

```{r}
ggplot(gapminder_2007) + geom_point(aes(x = gdpPercap, y = lifeExp,size = pop),
alpha=0.5) + scale_size_area(max_size = 10)
```

```{r}
gapminder_1957 <- gapminder %>% filter(year==1957) 
ggplot(gapminder_1957)+ 
aes(x=gdpPercap, y=lifeExp, color =continent, size =pop) +geom_point(alpha=0.7) + scale_size_area(max_size =15)
```

```{r}
gapminder_1957 <- gapminder %>% filter(year==2007| year == 1957) 
ggplot(gapminder_1957)+ 
aes(x=gdpPercap, y=lifeExp, color =continent, size =pop) +geom_point(alpha=0.7)+ scale_size_area(max_size =15) +facet_wrap(~year)
```

##### Bar Charts

```{r}
gapminder_top5 <- gapminder %>% 
  filter(year==2007) %>% 
  arrange(desc(pop)) %>% 
  top_n(5, pop)

gapminder_top5
```

```{r}
ggplot(gapminder_top5)+ aes(x= country, y=pop) + geom_col() 
ggplot(gapminder_top5)+ aes(x= country, y=lifeExp) + geom_col()
```

```{r}
ggplot(gapminder_top5)+ aes(x= country, y=pop, fill = continent) + geom_col()  
ggplot(gapminder_top5)+ aes(x= country, y=pop, fill = lifeExp) + geom_col() 
```

```{r}
ggplot(gapminder_top5)+ 
aes(x= reorder(country, -pop), y=pop, fill = gdpPercap)+ geom_col()  
ggplot(gapminder_top5)+ 
  aes(x= reorder(country, -pop), y=pop, fill = country) +
geom_col(col = "gray30")  +guides(fill = "none")

```

```{r}
head(USArrests) 
USArrests$State <- rownames(USArrests) 
ggplot(USArrests) +aes(x=reorder(State,Murder), y=Murder) +geom_point() +
geom_segment(aes(x=State, xend=State, y=0, yend=Murder), color="violet")+
coord_flip()
```

##### Combining Plots

```{r}
# install.packages ("patchwork")
library(patchwork) 
p1 <-  ggplot(gapminder_top5)+ 
  aes(x= reorder(country, -pop), y=pop, fill = gdpPercap) + geom_col()   

p2 <- ggplot(gapminder_top5)+ aes(x= reorder(country, -pop), y=pop, fill = country) + geom_col(col = "gray30")  + guides(fill = "none")

(p1 |p2)
```
