---
title: "Halloween Mini-Project"
author: "Adithi Kumar (PID: A16653979)"
format: pdf
editor: visual
---

# Importing Candy Data

```{r}
candy_file <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv"
candy = read.csv(candy_file, row.names = 1)

head(candy)
```

## **Q1**. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

There are 85 different types of candy in this dataset

## **Q2**. How many fruity candy types are in the dataset?

```{r}
sum(candy[,2]) 
```

There are 38 fruity candy types in the dataset.

# What is your favorite candy?

```{r}
candy["Twix", ]$winpercent
```

## **Q3**. What is your favorite candy in the dataset and what is it's `winpercent` value?

```{r}
candy["Fun Dip", ]$winpercent

```

My favorite candy in the dataset is probably fun dip; it has a winpercent value of 39.1855%.

## **Q4**. What is the `winpercent` value for "Kit Kat"?

```{r}
candy["Kit Kat", ]$winpercent
```

Kit Kats have a winpercent value of 76.7686%.

## **Q5**. What is the `winpercent` value for "Tootsie Roll Snack Bars"?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```

Tootsie Roll Snack Bars have a winpercent value of 49.6535.

```{r}
library("skimr") 
skim(candy)
```

## **Q6**. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

The winpercent variable looks like it's on a different scale to the majority of the other columns in the dataset.

## **Q7**. What do you think a zero and one represent for the `candy$chocolate` column?

A 0 means that particular candy is not contain chocolate while a 1 means that the candy DOES contain candy.

## **Q8**. Plot a histogram of `winpercent` values

```{r}
library(ggplot2) 
ggplot(candy, aes(x =candy$winpercent)) +
       geom_histogram() 
```

## **Q9**. Is the distribution of `winpercent` values symmetrical?

The distribution looks like it is slightly skewed right

## **Q10**. Is the center of the distribution above or below 50%?

The center of distribution is below 50%.

## **Q11**. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
# as.logical reurns the TRUE/FALSE values of a column #candy$winpercent will give us only the winpercent of the TRUE values

choc_winpercent <- candy$winpercent[as.logical(candy$chocolate)] 

mean(choc_winpercent)

fruity_winpercent <- candy$winpercent [as.logical(candy$fruity)] 

mean(fruity_winpercent)

```

On average, chocolate candy is higher ranked than fruity candy; chocolate candy has an average winpercent value of about 60.92 while fruity candy has an average winpercent value of about 44.12%.

## **Q12**. Is this difference statistically significant?

```{r}
t.test(choc_winpercent, fruity_winpercent)
```

Since the p-value of 2.871e-08 is less than 0.05, the difference between the average winpercent values between chocolate and fruity candy is statistically significant.

# Overall Candy Rankings

## **Q13**. What are the five least liked candy types in this set?

```{r}
# order will order dataset from least to most 
head(candy[order(candy$winpercent),], n=5)

```

The five least liked candy types are Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, and Jawbusters.

## **Q14**. What are the top 5 all time favorite candy types out of this set?

```{r}
tail(candy[order(candy$winpercent),], n=5)
```

The top five all time favorite candy tupes are Snickers, Kit Kat, Twix, Reese's Miniatures, Reese's Peanut Butter Cup.

## **Q15**. Make a first barplot of candy ranking based on `winpercent` values.

```{r}
library(ggplot2) 
ggplot(candy, aes(winpercent, rownames(candy)))+ 
         geom_col()
```

## **Q16**. This is quite ugly, use the `reorder()` function to get the bars sorted by `winpercent`?

```{r}
library(ggplot2)  

ggplot(candy,aes(winpercent, reorder(rownames(candy),winpercent)))+ 
         geom_col()
```

```{r}
#Add some color 
library(ggplot2)  

#vectors for all black values
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"


ggplot(candy,aes(winpercent, reorder(rownames(candy),winpercent), ))+ 
         geom_col(fill =my_cols) 

```

## **Q17**. What is the worst ranked chocolate candy?

The worst ranked chocolate candy is Sixlets.

## **Q18**. What is the best ranked fruity candy?

The worst ranked fruity candy is Nik L Nip.

# Taking a look at pricepercent

```{r}
library(ggrepel) 
# plot of price vs. win percent 
ggplot(candy) + aes (winpercent, pricepercent, label = rownames(candy)) + 
  geom_point(col =my_cols) +   
  geom_text_repel(col = my_cols, size =3.3, max.overlaps =53)
```

## **Q19**. Which candy type is the highest ranked in terms of `winpercent` for the least money - i.e. offers the most bang for your buck?

Based on the plot, the candy ranked highest in terms of winpercent for the least amount of money is Reeses Miniatures.

```{r}
ord <- order(candy$pricepercent, decreasing =TRUE) 
head (candy [ord, c(11,12)], n=5)  
tail (candy [ord, c(11,12)], n=5)  
```

## **Q20**. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

The top 5 most expensive candy types in the dataset are Nik L Nip, Nestle Smarties, Ring pop, Hershey's Krackel, and Hershey's Milk Chocolate. The least popular of these are Nik L Nip with a winpercent of 22.44534%.

# Exploring the Correlation Structure

```{r}
library(corrplot) 
cij <- cor(candy) 
corrplot(cij)
```

Corrplot looks at what variables are correlated and anti-correlated. Correlated means the two variables follow each other; when one goes up, the other will go up as well! Anti-correlated is a negative correlation where one goes up and the other goes down. Magnitude of correlation is represented by the radius of the circle: the larger the circle, the stronger the correlation.

## **Q22**. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

The variables chocolate and fruit are anti-correlated.

## **Q23**. Similarly, what two variables are most positively correlated?

The two variables that are most positively correlated are chocolate and winpercent.

# Principal Component Analysis

```{r}
pca <- prcomp(candy, scale =TRUE) 
summary (pca)
```

```{r}
plot(pca$x[,1:2])
```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}
#make new df with PCA and candy data  
my_data <- cbind (candy, pca$x [,1:3])

p <- ggplot(my_data) + 
  aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols) 
p
```

```{r}
library(ggrepel) 
p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 47)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")

```

```{r}
#use plotly for interactive labels :)
#library(plotly)
#ggplotly(p)
```

```{r}
par(mar = c(8,4,2,2)) 
barplot(pca$rotation[,1], las =2, ylab = "PC1 Contribution")
```

## **Q24**. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

The original variables that are picked up strongly by PC1 in the positive direction are fruity, hard, and pluribus. Yes! Each variable influences the direction of each PC with different magnitudes. The variables fruity, hard, and pluribus have variables that they are anti-correlated with that are picked up strongly by PC1 in the negative direction (chocolate and bar). If the candy is a bar, it is less likey to be pluribus(come in a form of multiple candies) or fruity. If a candy is chocolate, it is less likely to be fruity. PC1 finds dimensions that maximizes variance so variables that are strongly anti-correlated would be on the ends of the data; they would be the extremes that need to be picked up by PC1 for PC1 to include the most variance of the data. Variables that are correlated end up facing the same direction (positive or negative) on the plot; chocolate and winpercent are strongly correlated and face the negative direction.
